Efficient Calculation of Polynomial Features on Sparse Matrices

### 2016-02-18 paper/workshop/sparse_poly.tex
Much reduced section on formula derivations
Removed most algebra and put many displayed equations inline.

For the second degree case, we seek a map from matrix indices $(i, j)$ (with $i < j$ and $0 \le i < D$) to numbers $f(i, j)$ with $0 \le f(i, j) < \frac{D(D-1)}{2}$, one that follows the pattern indicated by 
\begin{align}
\begin{bmatrix}
x & 0 & 1 & 3 \\
x & x & 2 & 4 \\
x & x & x & 5 \\
x & x & x & x
\end{bmatrix}
\label{eq:4x4mat}
\end{align}
where the entry in row $i$, column $j$, displays the value $f(i, j)$.

To simplify slightly, we introduce a notation for the $n$th triangular number, 
\begin{align}
T_2(n) = \frac{n(n+1)}{2}
\end{align}
\noindent
The subscript $2$ indicates that these are triangles in two dimensions; we'll use $T_3(n)$ to indicate the $n$th tetrahedral number, and so on for higher dimensions.

Observe that in Equation~\ref{eq:4x4mat}, each entry in column $j$ (for $j > 0$) lies in the range
\begin{align}
T_2(j-1) &\le e < T_2(j).
\end{align}
\noindent
And in fact, the entry in the $i$th row of that column is just $i + T_2(j-1)$. Thus we have
\begin{align}
f(i, j) 
&= i + T_2(j-1)\\
&= i + \frac{(j-1)j}{2}\\
&=  \frac{2i + j^2-j}{2}.
\end{align}

For instance, in column $j = 2$ in our example (the \emph{third} column), the entries range from $1$ to $2$, while $T_2(j-1) = T_2(1) = 1$ and $T_2(j) = T_2(2) = 3$, and the entry in column $j = 2$, row $i = 1$ is 
$i + T_2(j-1) = 1 + 1 = 2$.

\subsubsection{Other indices}
With one-based indexing in both the domain and codomain, the formula above becomes
\begin{align}
f_1(i, j) &= 1+ f(i-1, j-1) \\
&= 1+ f(i-1, j-1) \\
& = \frac{2 + 2(i-1) + (j-1)^2-(j-1)}{2}\\
& = \frac{2i + j^2 - 3j + 2}{2}
\end{align}

\subsubsection{Polynomial Features}
For polynommial features, we seek a map from matrix indices $(i, j)$ (with $i \le j$ and $0 \le i < D$) to numbers $g(i, j)$ with $0 \le f(i, j) < \frac{D(D+1)}{2}$, one that follows the pattern indicated by 
\begin{align}
\begin{bmatrix}
 0 & 1 & 3 & 6 \\
 x & 2 & 4 & 7\\
 x & x & 5 & 8 \\
 x & x & x & 9
\end{bmatrix}
\label{eq:4x4mat-poly}
\end{align}
i.e., essentially the same task as before, except that the diagonal is included. One can regard all but the last column of entries in Equation~\ref{eq:4x4mat-poly} as corresponding to the entries in Equation~\ref{eq:4x4mat}, but shifted to the left. Thus the formula for $g(i, j)$ is simply the formula for $f$, shifted by 1, i.e., 
\begin{align}
g(i, j) &= f(i, j+1)  \\
&=  \frac{2i + (j+1)^2-(j+1)}{2}\\
&=  \frac{2i + j^2 + j + 1)}{2}.
\end{align}
Alternatively, we can write this as
\begin{align}
g(i, j) &= i + T_2(j),
\end{align}
\noindent
and get the same result.

\subsubsection{Higher dimensions}
To handle three-way interactions, we need to map triples of indices in a 3-index array to a flat list, and similarly for higher-order interactions.

For three indices, $i,j,k$, with $i < j < k$ and $0 \le i,j,k < D$, we have a similar recurrence. Calling the mapping $h$, we have 
\begin{align}
h(i,j,k) = i + T_2(j-1) + T_3(k-2);
\end{align}
if we define $T_1(i) = i$, then this has the very regular form
\begin{align}
h(i,j,k) =  T_1(i) + T_2(j-1) + T_3(k-2);
\end{align}
and from this, the generalization to higher dimensions is straightforward. The formulas for ``higher triangular numbers'', i.e., those defined by
\begin{align}
T_k(n) &= \sum_{i=1}^n T_{k-1}(n)
\end{align}
for $k > 1$ can be determined inductively. For $k = 3$, the result is 
\begin{align}
T_3(n) &= \sum_{i=1}^n T_{2}(n)\\
&= \frac{n^3 + 3n^2 + 2n}{6},
\end{align}
so that the formula for 3-way interactions, with zero-based indexing, becomes 
\begin{align}
h(i, j, k) &= 1 + (i-1) + \frac{(j-1)j}{2} + \\
& \frac{(k-2)^3 + 3(k-2)^2 + 2(k-2)}{6}. 
\end{align}
\subsubsection{Higher-dimension polynomial features}
For the case where we include the diagonal in higher dimensions, we must shift $j$ by $1$, $k$ by $2$, and so on, and the formula becomes
\begin{align}
\ell(i,j,k) =  T_1(i) + T_2(j) + T_3(k),
\end{align}
with analogous formulas for higher degree polynomial interactions.

@@

Calculating $k$-degree polynomial features via our method for a vector of dimensionality $D$ and density $d$ requires $dD$ choose $k$ (with repetition) products.
The big O of the algorithm is therefore given by
\begin{align}
O\left(\binom{dD+k-1}{k}\right) & = O\left(\frac{(dD+k-1)!}{k!(dD-1)!}\right)\\
& = O\left(\frac{(dD+k-1)(dD+k-2) \dots (dD)}{k!}\right)\\
& = O\left((dD+k-1)(dD+k-2) \dots (dD)\right) \mbox{ for } k \ll dD\\
& = O\left(d^kD^k\right)
\end{align}

>>>>

For the second degree case, we seek a map from matrix indices $(i, j)$ (with $0 \le i < j < D$ ) to numbers $f(i, j)$ with $0 \le f(i, j) < \frac{D(D-1)}{2}$, one that follows the pattern indicated by 
\begin{align}
\begin{bmatrix}
x & 0 & 1 & 3 \\
x & x & 2 & 4 \\
x & x & x & 5 \\
x & x & x & x
\end{bmatrix}
\label{eq:4x4mat}
\end{align}
where the entry in row $i$, column $j$, displays the value $f(i, j)$. We let $T_2(n) = \frac{1}{2} n(n+1)$ 
be the $n$th triangular number; then in Equation~\ref{eq:4x4mat}, column $j$ (for $j > 0$) contains entries with  
$T_2(j-1) \le e < T_2(j)$; the entry in the $i$th row is just $i + T_2(j-1)$. Thus we have
$
f(i, j) 
= i + T_2(j-1) =  \frac{1}{2}(2i + j^2-j).$
For instance, in column $j = 2$ in our example (the \emph{third} column), the entry in row $i = 1$ is 
$i + T_2(j-1) = 1 + 1 = 2$.

With one-based indexing in both the domain and codomain, the formula above becomes
$f_1(i, j)  = \frac{1}{2}(2i + j^2 - 3j + 2).$

For \emph{polynomial} features, we seek a similar map $g$, one that also handles the case $i = j$. In this case, a similar analysis yields
$ g(i, j) = i + T_2(j) = \frac{1}{2} (2i + j^2 + j + 1).$

To handle \emph{three-way interactions}, we need to map triples of indices in a 3-index array to a flat list, and similarly for higher-order interactions. For this, we'll need the tetrahedral numbers $T_3(n) = \sum_{i=1}^n T_{2}(n) = 
\frac{1}{6}(n^3 + 3n^2 + 2n)$.

For three indices, $i,j,k$, with $0 \le i < j < k < D$, we have a similar recurrence. Calling the mapping $h$, we have 
\begin{align}
h(i,j,k) = i + T_2(j-1) + T_3(k-2);
\end{align}
if we define $T_1(i) = i$, then this has the very regular form
\begin{align}
h(i,j,k) =  T_1(i) + T_2(j-1) + T_3(k-2);
\end{align}
and from this the generalization to higher dimensions is straightforward. The formulas for ``higher triangular numbers'', i.e., those defined by
\begin{align}
T_k(n) &= \sum_{i=1}^n T_{k-1}(n)
\end{align}
for $k > 1$ can be determined inductively.

The explicit formula for 3-way interactions, with zero-based indexing, is 
\begin{align}
h(i, j, k) &= 1 + (i-1) + \frac{(j-1)j}{2} + \\
& \frac{(k-2)^3 + 3(k-2)^2 + 2(k-2)}{6}. 
\end{align}

@@

Calculating $k$-degree polynomial features via our method for a vector of dimensionality $D$ and density $d$ requires $\binom{dD}{k}$ (with repetition) products.
The complexity of the algorithm, for fixed $k \ll dD$, is therefore
\begin{align}
O\left(\binom{dD+k-1}{k}\right) & = O\left(\frac{(dD+k-1)!}{k!(dD-1)!}\right)\\
& = O\left(\frac{(dD+k-1)(dD+k-2) \dots (dD)}{k!}\right)\\
& = O\left((dD+k-1)(dD+k-2) \dots (dD)\right) \mbox{ for } k \ll dD\\
& = O\left(d^kD^k\right)
\end{align}
